---
title: "My project"
format: html
editor: visual
---

# Description

REASON FOR CHOOSING THE DATA

I chose the data about the composition of solid waste because the issue of solid waste management is a pressing challenge faced by the entire world.

As the global population continues to urbanize and consumption patterns shift, the generation of solid waste is expected to increase and understanding the composition is a crucial first step in handling the issue.

```{r}
library(tidyverse)
```

# Import

```{r}
readxl::read_excel(path = here::here("data/raw/Dominic.xlsx"))
```

```{r}
Dominic <- readxl::read_xlsx(path = "/cloud/project/data/raw/Dominic.xlsx")
```

```{r}
dominic_processed <- Dominic |> 
  select(iso3c:composition_plastic_percent) |> 
  rename(region = region_id,
         country = country_name,
         population = population_population_number_of_people,
         food_organic_waste = composition_food_organic_waste_percent,
         glass = composition_glass_percent,
         metal = composition_metal_percent,
         other = composition_other_percent,
         paper_cardboard = composition_paper_cardboard_percent,
         plastic = composition_plastic_percent) |> 
   filter(food_organic_waste != "NA",
         glass != "NA",
         metal != "NA",
         other != "NA",
         paper_cardboard != "NA",
         plastic != "NA")

write_csv(dominic_processed, file = "/cloud/project/data/processed/dominic_processed.csv")

```
